$Z_3$ topological order in the quantum dimer-pentamer model

### 2017-02-26 qdpm_paper.tex
fixed S_d axis (real and positive) and removed note from tex file.
Minor changes
minor fixes to tex and peak_scalling figure

\begin{abstract}
We introduce the quantum dimer-pentamer model (QDPM) on the square lattice. This model is a generalization of the square lattice quantum dimer model (QDM) as its configuration space comprises fully-packed hard-core dimer coverings as well as dimer configurations containing pentamers, where four dimers touch a vertex. Thus in the QDPM, the fully-packed, hard-core constraint of the QDM is relaxed such that the local dimer number at each vertex is fixed modulo 3; correspondingly, the local $U(1)$ gauge symmetry of the QDM Hilbert space is reduced to a local $Z_3$ gauge symmetry in the QDPM. We construct a local Hamiltonian for which the Rokhsar-Kivelson (RK) equal superposition state is the exact ground state and has a 9-fold topological degeneracy on the torus. Using Monte Carlo calculations, we find no spontaneous symmetry breaking in the RK wavefunction and that its dimer-dimer correlation function decays exponentially. By doping the QDPM RK state with a pair of monomers, we demonstrate that $Z_3$ electric charges are deconfined. Additionally, we introduce a $Z_3$ magnetic string operator which shows no signatures of magnetic vortex condensation also decay exponentially. These numerical results suggest that the ground state of the QDPM is a dimer liquid with $Z_3$ topological order.
\end{abstract}

@@

Strongly interacting quantum many-body systems are known to support a variety of exotic quantum phases of matter with no classical analogue. One class of such quantum phases of matter are topologically ordered quantum liquids; their non-trivial topological order can not be identified by a local order parameter as these phases of matter don't break conventional symmetries~\cite{Wen1990,Nayak2008}. Both because of of the exotic nature of topological phases and their potential use for quantum information processing~\cite{Freedman2001,Kitaev2003}, there is great interest in the development of theoretical models which display these phases and potentially could provide candidate experimental systems in which to constructively engineer topological order~\cite{Duan2003,Jaksch2005,Lewenstein2007,Jiang2008d,Weimer2010a,Herdman2010c,Fowler2012a}.

@@

While on non-bipartite two-dimensionsal lattices, the QDM has a topologically ordered ground state~\cite{Moessner2001a,Fendley2002}, on the square lattice there is only a symmetric liquid ground state at an isolated critical quantum point~\cite{Leung1996,Syljuasen2006}. If the hard-core constraint of the QDM is relaxed such that only the parity of the number of dimer touching each vertex is fixed, the simplest non-trivial dynamics leads to a $Z_2$ topologically ordered ground state~\cite{Kitaev2003,Wen2003}. The existence of topologically ordered ground states in these models is intimately related to the local constraint which imposes a local gauge symmetry in the Hilbert space; the topological order that arises is directly related to a discrete gauge theory with the appropriate local gauge constraint~\cite{Moessner2001}.

Given that these models with dimer degrees of freedom and local constraints posses the simplest non-trivial local Hilbert space at each link, they provide a potentially simpler route to engineering exotic phases. The local constraints can be enforced by a local potential energy penalty and the minimal dynamics are often arise at the lowest order non-trivial order in perturbation theory from a related spin model. We will refer to such models with dimer degrees of freedom on links and local constraints as \emph{generalized quantum dimer modes} (GQDMs). Previous work has demonstrated that GQDMs on the square lattice support gapped $Z_2$ topologically ordered phases as well as gapless quantum critical points. On non-bipartite lattices, GQDMs support $Z_2$ topological order~\cite{Moessner2001a,Misguich2002} as well as doubled-semion phases~\cite{Qi2014,Buerschaper2014a}. A full characterization of all exotic phases that can exist within this framework is thus desirable.

@@

Additionally we present a numerical study of the ground state of the QDPM at an exactly soluble point. Using Monte Carlo calculations, we explicitly demonstrate the absence of symmetry breaking order and the exponential decay of correlation functions. Additionally, we define a $Z_3$ magnetic string operator, whose correlations display no evidence of magnetic vortex condensation. Finally, by a studying systems doped with a pair of  monomer defects, we demonstrate that monomers are deconfined in the QDPM. Thus we argue that the ground state of the QDPM is a $Z_3$ topologically ordered dimer liquid.

This paper is organized as follows. In \secref{sec:Background} we provide relevant background for the QDM and related models. In \secref{sec:QDPM} we present the details of the QPDM. \secref{sec:numerics} provides the results of a Monte Carlo study of the ground state of the QDPM at its exactly soluble point. Finally, \secref{sec:discussion} discusses potential future studies of the QDPM.

@@

The Hamiltonian of a GQDM is of the form
\begin{equation}
H_{\rm{GQDM}} = -\sum_\nu t_\nu  {T}_{\nu} + \sum_\nu v_\nu  {V}_{\nu} \label{eq:HGQDM}
\end{equation}
where ${T}_\nu$ are off-diagonal kinetic energy operators, ${V}_\nu$ are diagonal potential energy operators, $t_\nu$ and $v_\nu$ parametrize the strength of these terms, and each $\nu$ and represents a set of  neighboring links. We will limit our discussion to $t_\nu >0$. ${T}_\nu$ is constructed to generate resonances between two distinct allowed dimerazations of $\nu$, $c_\nu$ and its compliment $\bar{c}_\nu$, and is written as:
\begin{equation}
 {T}_{\nu} = \eket{c_\nu}\ebra{\bar{c}_\nu} + h.c.;
\end{equation}
in the simplest case, $ {T_\nu}$ involves the minimum non-trivial dimer dynamics that preserve the constraint. Similarly, the potential energy term provides an energy cost for states with these local configurations:
\begin{equation}
 {V}_{\nu} = \eket{c_\nu}\ebra{c_\nu} +  \eket{\bar{c}_\nu}\ebra{\bar{c}_\nu}.
\end{equation}
Therefore we may define a Hamiltonian for a GQDM using \Eqref{eq:HGQDM} and defining a set of dimer resonances $\{(c_\nu,\bar{c}_\nu)\}$ and their corresponding strengths $\{(t_\nu,v_\nu)\}$:
\begin{equation}
\HGQDM \biggl( \bigl\{\left(c_\nu,\bar{c}_\nu\right)\},\{\left(t_\nu,v_\nu\right)\bigr\}\biggr).
\end{equation}
 In general the $[T_\nu, T_{\nu'}] \neq 0 $ and $[T_\nu,V_{\nu'}]\neq0$ when $\nu$ and $\nu'$ share links, and thus the ground state of $H_{\rm{GQDM}}$ is not explicitly known.

However, at the special points in the $\{(t_\nu,v_\nu)\}$ parameter space, known as the Rokhsar-Kivelson (RK) points, when \Eqref{eq:HGQDM} can be written as the sum of projection operators
\begin{equation}
H_{\rm{RK}} = \sum_\nu  {h}_\nu \label{eq:HRK},
\end{equation}
where all $h_\nu$ are projectors with eigenvalues $\{0,1\}$, the exact ground state is known to be the zero energy state of \Eqref{eq:HRK} that is simultaneously annihilated by all ${h}_\nu$. Such a zero energy state, known as the RK state, may be formed from an equal superposition of all dimerizations that are connected by $\{T_\nu\}$:
\begin{equation}
\eket{\Phi_{\RK}^{\Omega}} = \sum_{C \in \Omega} \eket{C}
\end{equation} 
where $\Omega$ is a subset of the Hilbert space connected by $\{T_\nu\}$. Thus $\{ \ket{\Phi_{\RK}^\Omega} \}$ are the exact, degenerate ground states of $H_{\rm{GQDM}}$ with a degeneracy labelled by $\Omega$ at the RK point.

Because of the local constraints, all states $\ket{\psi}$ in the Hilbert space of a GQDM are invariant under local gauge transformations $G_v \ket{\psi} = \ket{\psi}$ a the vertices, $G_v$ of the general form~\cite{Moessner2001}
\begin{equation}
G_v = e^{i \alpha_v Q_v}, \label{eq:Gv}
\end{equation}
where the allowed values of $\alpha_v$ depend on the constraint.  
By construction, the Hamiltonian is gauge invariant and thus commutes with $G_v$
\begin{equation}
\left[ H_{\rm{GQDM}}, G_v \right] = 0. \notag
\end{equation}

@@

The gauge symmetry allows for the construction of a gauge invariant magnetic string operator $S_\gamma$ along an oriented path $\gamma$ path on the dual lattice. We first orient all the links of the square lattice such that they point from the $A$ sublattice to the $B$ sublattice (see \figref{fig:orientedSL}) and define an oriented dimer flux $\Phi_\gamma$ through the path:
\begin{equation}
\Phi_\gamma \equiv \sum_{\ell \in \gamma} \rm{sgn}\left(\ell,\gamma\right) n_\ell \label{eq:phi}
\end{equation}
where $\mathrm{sgn}(\ell,\gamma)$ is positive (negative) if $\ell$ is oriented to the right with respect to the orientation of $\gamma$. Additionally, we define background flux $\tilde{\Phi}_\ell$
\begin{equation}
\tilde{\Phi}_\gamma \equiv  \sum_{\ell \in \gamma} \rm{sgn}\left(\ell,\gamma\right) \tilde{n}_\ell \label{eq:phiBackground}
\end{equation}
where $n_\ell$ is the number operator for a static background dimer configuration which is chosen to fix the gauge. With this gauge fixing, the total charge enclosed in a closed loop $\Gamma$ is $Q_\Gamma = \Phi_\Gamma- \tilde{\Phi}_\Gamma$. An alternative physical picture is that there are static charges of charge $+c_0$ ($-c_0$) sitting at the vertices of the $A$($B$) sub-lattice, which are the source of the non-vanishing $\Phi_\Gamma$ flux through a closed loop.

@@

The two unique fixed-number constraints on the square lattice are lead to QDM in the odd sector ($n_v = 1$) and the quantum full-packed loop model (QFPLM) in the even sector ($n_v = 2$). For configurations that satisfy these a number constraints, $Q_v$ strictly vanishes, and thus all states in the Hilbert space are invariant under $G_v$ for $0 \leq \alpha_v \leq 2 \pi$ in \Eqref{eq:Gv}; this represents a local $U(1)$ gauge symmetry~\cite{Moessner2001}.

The Hamiltonian for these models gives dynamics to parallel dimers on the plaquettes $\bplqt$ of the lattice:
\begin{equation}
\HQDM\left(t,v\right) \equiv \HGQDM \biggl( \bigl\{ \left(\vrtplqt,\hrzplqt\right) \bigr\},\bigl\{ \left(t,v\right) \bigr\} \biggr)  \label{eq:HQDM}
\end{equation}
where we have chosen $t_\bplqt = t$, $v_\bplqt =v, \forall \bplqt$.
%\begin{align}
%    \label{}
%   \HQDM = &-t\sum_{\bplqt} \biggl( \eket{\vrtplqt} \ebra{\hrzplqt} +h.c. \biggr) \notag\\
%    					&+v\sum_{\bplqt} \biggl( \eket{\vrtplqt} \ebra{\vrtplqt} +\eket{\hrzplqt} \ebra{\hrzplqt} \biggr). \label{eq:HQDM}
%\end{align}
In this case of a fixed number constraint, the Hamiltonian conserves $Q_v$ and therefore the dimer flux through any closed loop such that:
\begin{equation}
\left[ \Phi_\Gamma, \HQDM \right] = 0
\end{equation}
and therefore we can define two conserved topological winding numbers $\Phi_{X}$ \& $\Phi_{Y}$ which are the dimer flux through topologically non-trivial loops that wind around each axis of the lattice when the system has periodic boundary conditions (see \figref{fig:orientedSL}). On an lattice of dimensions $L_x\times L_y$, the eigenvalues of $\Phi_X$ and $\Phi_Y$ are integers in the range
\begin{equation}
-L_{x}/2 \leq  \Phi_{X} \leq L_{x}/2,
\end{equation}
and similarly for $\Phi_Y$, thus there are $\mathcal{O} (L_x L_y)$ distinct topological sectors.

@@

Defects in these models that violate the local dimer number constraint and thus carry non-zero U(1) electric charge. \tabref{tab:defects} lists all the possible defects and their associated U(1) charges in QDM and QFPLM. When a defect lives on sublattice A, it has the opposite charge that the same defect on sublattice B has. Note that in the QDM, the maximum charge is $\pm 3$, and there are six distinct possible non-trivial electric charges of defects.

@@

We now relax the local constraint to a parity constraint such that only $Q_v^{Z_2} \equiv (Q_v \mod 2)$ is required to vanish; the resulting GQDM are the even and odd sectors are the  even \& odd Ising gauge theories (IGT) with constraints $n_v \in \{0,2,4\}$ and $n_v \in \{1,3\}$, respectively. In these cases, the local constraint requires $\alpha_v \in \{ 0, \pi\}$ in \Eqref{eq:Gv} and therefore the gauge symmetry is reduced from $U(1)$ to $Z_2$. The set of allowed dimer resonances $\mathcal{C}_{\rm{IGT}}$ is expanded accordingly:
\begin{align}
\mathcal{C}_{\rm{IGT}} &\equiv \bigl\{ \left(c_{\Box,i},\bar{c}_{\Box,i}\right) \bigr\} \\
&=\bigl\{
\left(\vrtplqt,\hrzplqt\right),\left(\threeleftplqt,\oneleftplqt\right),\left(\twodownplqt,\twoupplqt\right),\left(\bplqt,\fplqt\right) \bigr\}
\end{align}
where we implicitly include symmetry related resonances. We will limit or discussion to $v_\bplqt=0$ such that the corresponding Hamiltonian is
\begin{equation}
\HIGT\left(t\right) \equiv \HGQDM \biggl( \mathcal{C}_{\rm{IGT}},\bigl\{ \left(t,0\right) \bigr\} \biggr).
\end{equation}
In this case the equal superposition RK wavefunction is the exact ground state occurs in the absence of a potential energy term, as we can define a projection operator on each plaquette
\begin{equation}
h_\Box \equiv \frac{1}{2} \left( \mathbb{1} - \sum_i \eket{c_{\Box,i}}\ebra{\bar{c}_{\Box,i}} + h.c. \right)
\end{equation}
such that $\HIGT$ takes the form of \Eqref{eq:HRK} (up to a constant), and the RK state is the ground state for all values of $t$.

@@

We may understand the emergence of the $Z_2$ topologically ordered phase of IGTs from the QDM and QFPLM from the following picture. The additional vertex configurations that are allowed in the IGTs relative to the models with a dimer number constraint  have charge U(1) charge $Q_v = \pm 2$ (see \tabref{tab:defects}). It is known that coupling a $U(1)$ gauge field to a dynamics charge $q$ matter field can result in an $Z_q$ gauge theory~\cite{Fradkin1979}; as in the IGT's we have added dynamic charge 2 defects to the underlying $U(1)$ symmetric model, and we may expect a $Z_2$ gauge theory to result. We note that pentamers are charge $\pm3$ objects in the QDM (see \tabref{tab:defects}); thus below we investigate whether introducing pentamers into the QDM leads to an effective $Z_3$ gauge theory (see \appref{sec:Z3GT} for a review of $Z_3$ gauge theory).

@@

We will now present a numerical study of the dimer density correlations in the RK ground state of the QDPM. In \figref{fig:dimer_3way} (a) we plot the bare dimer-dimer density correlation function $\langle n_{\bm{0}} n_{\bm{r}}\rangle$ on a $L=32$ lattice; here the origin $\bm{0}$ represents the orange link, and $\bm{r}$ varies over all other links of the lattice. These bare dimer correlations display only short range correlations no obvious symmetry breaking. In \figref{fig:dimer_3way} (b) we show the equal-time dimer-dimer structure factor, 
\begin{equation}
S_d\left( \bm{k} \right) \equiv \frac{1}{ L^4} \sum_{i,j} e^{-i \bm{k} \cdot \left(\bm{r}_i - \bm{r}_j\right)} \left \langle n_{\bm{r}_i} n_{\bm{r}_j} \right \rangle
\end{equation}
where we restrict the sum to horizontal links, for a $L=64$ lattice. 
\figref{fig:dimer_3way} (c) shows the magnitude \note{(Owen: Is $S_d$ real \& positive?)} of $S_d(\bm{k})$ for a particular path through the Brillouin zone. The dimer structure factor displays broad maxima at $\bm{k} = (0, \pm \pi)$ and $\bm{k} = (\pm \pi, \pm \pi)$. The lack of rotational symmetry of $S_d(\bm{k})$ is due to the fact that we have isolated horizontal dimers; the correlations between vertical dimers display the same behavior, rotated by $\pi/2$.
We have plotted the finite size scaling of the intensity of the broad maximum of $S_d$ in  \figref{fig:bragg_scaling}, which we find to vanish in the thermodynamic limit; these maxima are therefore consistent with finite size effects due to short distance correlations. In \appref{sec:dimer_order_parameters} we additionally have plotted order parameters that specifically test for rotational and translational symmetry breaking; in all cases we find no symmetry breaking in the thermodynamic limit. Thus we see no evidence of symmetry breaking in the thermodynamic limit.

@@

Additionally we have studied the nature of the decay of the connected dimer-dimer correlation function,
  \begin{equation}
    C_{\mathrm{d}} \left(r\right) \equiv \langle n_{\bm{0}} n_{\bm{r}} \rangle - \langle n_{\bm{0}} \rangle   \langle n_{\bm{r}} \rangle   
\end{equation}
where $r$ is the distance between the links $\bm{0}$ and $\bm{r}$. In \figref{fig:spatial_dmr_cor}, $C_d$ between parallel links is plotted as a function of separation, for $\bm{r}$ perpendicular to the orientation of the dimers for a $L=64$ lattice. We find that $C_d$ decays exponentially with distance having a correlations length  $\chi\approx 1$. Additionally, $C_d$ displays a sub-lattice oscillation that is consistent with the finite-size maximum in $S_d$ at $(0,\pi)$. Thus we find that the RK point of the QDPM is a symmetric dimer liquid with exponentially decaying dimer correlations, in contrast with the power-law decaying correlations of the QDM.

>>>>

\begin{abstract}
We introduce the quantum dimer-pentamer model (QDPM) on the square lattice. This model is a generalization of the square lattice quantum dimer model (QDM) as its configuration space comprises fully-packed hard-core dimer coverings as well as dimer configurations containing pentamers, where four dimers touch a vertex. Thus in the QDPM, the fully-packed, hard-core constraint of the QDM is relaxed such that the local dimer number at each vertex is fixed modulo 3; correspondingly, the local $U(1)$ gauge symmetry of the QDM Hilbert space is reduced to a local $Z_3$ gauge symmetry in the QDPM. We construct a local Hamiltonian for which the Rokhsar-Kivelson (RK) equal superposition state is the exact ground state and has a 9-fold topological degeneracy on the torus. Using Monte Carlo calculations, we find no spontaneous symmetry breaking in the RK wavefunction and that its dimer-dimer correlation function decays exponentially. By doping the QDPM RK state with a pair of monomers, we demonstrate that $Z_3$ electric charges are deconfined. Additionally, we introduce a $Z_3$ magnetic string operator which shows no signatures of magnetic vortex condensation and with correlations that decay exponentially. These numerical results suggest that the ground state of the QDPM is a dimer liquid with $Z_3$ topological order.
\end{abstract}

@@

Strongly interacting quantum many-body systems are known to support a variety of exotic quantum phases of matter with no classical analogue. One class of such quantum phases of matter are topologically ordered quantum liquids; their non-trivial topological order can not be identified by a local order parameter as these phases of matter don't break conventional symmetries~\cite{Wen1990,Nayak2008}. Both because of of the exotic nature of topological phases and their potential use for quantum information processing~\cite{Freedman2001,Kitaev2003}, there is great interest in the development of theoretical models which display these phases and potentially could provide candidate experimental systems to constructively engineer topological order~\cite{Duan2003,Jaksch2005,Lewenstein2007,Jiang2008d,Weimer2010a,Herdman2010c,Fowler2012a}.

@@

While on non-bipartite two-dimensionsal lattices, the QDM has a topologically ordered ground state~\cite{Moessner2001a,Fendley2002}, on the square lattice there is only a symmetric liquid ground state at an isolated critical quantum point~\cite{Leung1996,Syljuasen2006}. If the hard-core constraint of the square lattice QDM is relaxed such that only the parity of the number of dimer touching each vertex is fixed, the simplest non-trivial dynamics leads to a $Z_2$ topologically ordered ground state~\cite{Kitaev2003,Wen2003}. The existence of topologically ordered ground states in these models is intimately related to the local constraint which imposes a local gauge symmetry in the Hilbert space; the topological order that arises is directly related to a discrete gauge theory with the appropriate local gauge constraint~\cite{Moessner2001}.

Given that these models with dimer degrees of freedom and local constraints posses the simplest non-trivial local Hilbert space at each link, they provide a potentially simpler route to engineering exotic phases. The local constraints can be enforced by a local potential energy penalty and the minimal dynamics often arise at the lowest non-trivial order in perturbation theory from a related spin model. We will refer to such models with dimer degrees of freedom on links and local constraints as \emph{generalized quantum dimer modes} (GQDMs). Previous work has demonstrated that GQDMs on the square lattice support gapped $Z_2$ topologically ordered phases as well as gapless quantum critical points. On non-bipartite lattices, GQDMs support $Z_2$ topological order~\cite{Moessner2001a,Misguich2002} as well as doubled-semion phases~\cite{Qi2014,Buerschaper2014a}. A full characterization of all exotic phases that can exist within this framework is thus desirable.

@@

Additionally, we present a numerical study of the ground state of the QDPM at an exactly soluble point. Using Monte Carlo calculations, we explicitly demonstrate the absence of symmetry breaking order and the exponential decay of correlation functions. Additionally, we define a $Z_3$ magnetic string operator, whose correlations display no evidence of magnetic vortex condensation. Finally, by a studying systems doped with a pair of  monomer defects, we demonstrate that monomers are deconfined in the QDPM. Thus we argue that the ground state of the QDPM is a $Z_3$ topologically ordered dimer liquid.

This paper is organized as follows. In \secref{sec:Background} we provide relevant background for the QDM and related models. In \secref{sec:QDPM} we present the details of the QDPM. \secref{sec:numerics} provides the results of a Monte Carlo study of the ground state of the QDPM at its exactly soluble point. Finally, \secref{sec:discussion} discusses potential future studies of the QDPM.

@@

The Hamiltonian of a GQDM is of the form
\begin{equation}
H_{\rm{GQDM}} = -\sum_\nu t_\nu  {T}_{\nu} + \sum_\nu v_\nu  {V}_{\nu} \label{eq:HGQDM}
\end{equation}
where ${T}_\nu$ are off-diagonal kinetic energy operators, ${V}_\nu$ are diagonal potential energy operators, $t_\nu$ and $v_\nu$ parametrize the strength of these terms, and each $\nu$ and represents a set of  neighboring links. We will limit our discussion to $t_\nu >0$. ${T}_\nu$ is constructed to generate resonances between two distinct allowed dimerazations of $\nu$, $c_\nu$ and its compliment $\bar{c}_\nu$, and is written as:
\begin{equation}
 {T}_{\nu} = \eket{c_\nu}\ebra{\bar{c}_\nu} + h.c.;
\end{equation}
in the simplest case, $ {T_\nu}$ involves the minimum non-trivial dimer dynamics that preserve the constraint. Similarly, the potential energy term provides an energy cost for states with these local configurations:
\begin{equation}
 {V}_{\nu} = \eket{c_\nu}\ebra{c_\nu} +  \eket{\bar{c}_\nu}\ebra{\bar{c}_\nu}.
\end{equation}
Therefore we may define a Hamiltonian for a GQDM using \Eqref{eq:HGQDM} and defining a set of dimer resonances $\{(c_\nu,\bar{c}_\nu)\}$ and their corresponding strengths $\{(t_\nu,v_\nu)\}$:
\begin{equation}
\HGQDM \biggl( \bigl\{\left(c_\nu,\bar{c}_\nu\right)\},\{\left(t_\nu,v_\nu\right)\bigr\}\biggr).
\end{equation}
 In general, $[T_\nu, T_{\nu'}] \neq 0 $ and $[T_\nu,V_{\nu'}]\neq0$ when $\nu$ and $\nu'$ share links, and thus the ground state of $H_{\rm{GQDM}}$ is not explicitly known.

However, at the special points in the $\{(t_\nu,v_\nu)\}$ parameter space, known as the Rokhsar-Kivelson (RK) points, when \Eqref{eq:HGQDM} can be written as the sum of projection operators
\begin{equation}
H_{\rm{RK}} = \sum_\nu  {h}_\nu \label{eq:HRK},
\end{equation}
where all $h_\nu$ are projectors with eigenvalues $\{0,1\}$, the exact ground state is known to be
the zero energy state of \Eqref{eq:HRK} that is simultaneously annihilated by all ${\bar{h}}_\nu$. Such a zero energy state, known as the RK state, may be formed from an equal superposition of all dimerizations that are connected by $\{T_\nu\}$:
\begin{equation}
\eket{\Phi_{\RK}^{\Omega}} = \sum_{C \in \Omega} \eket{C}
\end{equation} 
where $\Omega$ is a subset of the Hilbert space connected by $\{T_\nu\}$. Thus $\{ \ket{\Phi_{\RK}^\Omega} \}$ are the exact, degenerate ground states of $H_{\rm{GQDM}}$ with a degeneracy labelled by $\Omega$ at the RK point.

Because of the local constraints, all states $\ket{\psi}$ in the Hilbert space of a GQDM are invariant under local gauge transformations $G_v \ket{\psi} = \ket{\psi}$ the vertices, $G_v$ of the general form~\cite{Moessner2001}
\begin{equation}
G_v = e^{i \alpha_v Q_v}, \label{eq:Gv}
\end{equation}
where the allowed values of $\alpha_v$ depend on the constraint.  
By construction, the Hamiltonian is gauge invariant and thus commutes with $G_v$
\begin{equation}
\left[ H_{\rm{GQDM}}, G_v \right] = 0. \notag
\end{equation}

@@

The gauge symmetry allows for the construction of a gauge invariant magnetic string operator $S_\gamma$ along an oriented path $\gamma$ path on the dual lattice. We first orient all the links of the square lattice such that they point from the $A$ sublattice to the $B$ sublattice (see \figref{fig:orientedSL}) and define an oriented dimer flux $\Phi_\gamma$ through the path:
\begin{equation}
\Phi_\gamma \equiv \sum_{\ell \in \gamma} \rm{sgn}\left(\ell,\gamma\right) n_\ell \label{eq:phi}
\end{equation}
where $\mathrm{sgn}(\ell,\gamma)$ is positive (negative) if $\ell$ is oriented to the right with respect to the orientation of $\gamma$. Additionally, we define background flux $\tilde{\Phi}_\ell$
\begin{equation}
\tilde{\Phi}_\gamma \equiv  \sum_{\ell \in \gamma} \rm{sgn}\left(\ell,\gamma\right) \tilde{n}_\ell \label{eq:phiBackground}
\end{equation}
where $\tilde{n}_\ell$ is the number operator for a static background dimer configuration which is chosen to fix the gauge. With this gauge fixing, the total charge enclosed in a closed loop $\Gamma$ is $Q_\Gamma = \Phi_\Gamma- \tilde{\Phi}_\Gamma$. An alternative physical picture is that there are static charges of charge $+c_0$ ($-c_0$) sitting at the vertices of the $A$($B$) sub-lattice, which are the source of the non-vanishing $\Phi_\Gamma$ flux through a closed loop.

@@

The two unique fixed-number constraints on the square lattice lead to QDM in the odd sector ($n_v = 1$) and the quantum full-packed loop model (QFPLM) in the even sector ($n_v = 2$). For configurations that satisfy these number constraints, $Q_v$ strictly vanishes, and thus all states in the Hilbert space are invariant under $G_v$ for $0 \leq \alpha_v \leq 2 \pi$ in \Eqref{eq:Gv}; this represents a local $U(1)$ gauge symmetry~\cite{Moessner2001}.

The Hamiltonian for these models gives dynamics to parallel dimers on the plaquettes $\bplqt$ of the lattice:
\begin{equation}
\HQDM\left(t,v\right) \equiv \HGQDM \biggl( \bigl\{ \left(\vrtplqt,\hrzplqt\right) \bigr\},\bigl\{ \left(t,v\right) \bigr\} \biggr)  \label{eq:HQDM}
\end{equation}
where we have chosen $t_\bplqt = t$, $v_\bplqt =v, \forall \bplqt$.
%\begin{align}
%    \label{}
%   \HQDM = &-t\sum_{\bplqt} \biggl( \eket{\vrtplqt} \ebra{\hrzplqt} +h.c. \biggr) \notag\\
%    					&+v\sum_{\bplqt} \biggl( \eket{\vrtplqt} \ebra{\vrtplqt} +\eket{\hrzplqt} \ebra{\hrzplqt} \biggr). \label{eq:HQDM}
%\end{align}
In this case of a fixed number constraint, the Hamiltonian conserves $Q_v$ and therefore the dimer flux through any closed loop such that:
\begin{equation}
\left[ \Phi_\Gamma, \HQDM \right] = 0,
\end{equation}
We can define two conserved topological winding numbers $\Phi_{X}$ \& $\Phi_{Y}$ which are the dimer flux through topologically non-trivial loops that wind around each axis of the lattice when the system has periodic boundary conditions (see \figref{fig:orientedSL}). On a lattice of dimensions $L_x\times L_y$, the eigenvalues of $\Phi_X$ and $\Phi_Y$ are integers in the range
\begin{equation}
-L_{x}/2 \leq  \Phi_{X} \leq L_{x}/2,
\end{equation}
and similarly for $\Phi_Y$, thus there are $\mathcal{O} (L_x L_y)$ distinct topological sectors.

@@

Defects in these models violate the local dimer number constraint and thus carry non-zero U(1) electric charge. \tabref{tab:defects} lists all the possible defects and their associated U(1) charges in QDM and QFPLM. When a defect lives on sublattice A, it has the opposite charge that the same defect on sublattice B has. Note that in the QDM, the maximum charge is $\pm 3$, and there are six distinct possible non-trivial electric charges of defects.

@@

We now relax the local constraint to a parity constraint such that only $Q_v^{Z_2} \equiv (Q_v \mod 2)$ is required to vanish; the resulting GQDM are the even and odd sectors are the  even \& odd Ising gauge theories (IGT) with constraints $n_v \in \{0,2,4\}$ and $n_v \in \{1,3\}$, respectively. In these cases, the local constraint requires $\alpha_v \in \{ 0, \pi\}$ in \Eqref{eq:Gv} and therefore the gauge symmetry is reduced from $U(1)$ to $Z_2$. The set of allowed dimer resonances $\mathcal{C}_{\rm{IGT}}$ is expanded accordingly:
\begin{align}
\mathcal{C}_{\rm{IGT}} &\equiv \bigl\{ \left(c_{\Box,i},\bar{c}_{\Box,i}\right) \bigr\} \\
&=\bigl\{
\left(\vrtplqt,\hrzplqt\right),\left(\threeleftplqt,\oneleftplqt\right),\left(\twodownplqt,\twoupplqt\right),\left(\bplqt,\fplqt\right) \bigr\}
\end{align}
where we implicitly include symmetry related resonances. We will limit our discussion to $v_\bplqt=0$ such that the corresponding Hamiltonian is
\begin{equation}
\HIGT\left(t\right) \equiv \HGQDM \biggl( \mathcal{C}_{\rm{IGT}},\bigl\{ \left(t,0\right) \bigr\} \biggr).
\end{equation}
In this case the equal superposition RK wavefunction is the exact ground state and occurs in the absence of a potential energy term, as we can define a projection operator on each plaquette
\begin{equation}
h_\Box \equiv \frac{1}{2} \left( \mathbb{1} - \sum_i \eket{c_{\Box,i}}\ebra{\bar{c}_{\Box,i}} + h.c. \right)
\end{equation}
such that $\HIGT$ takes the form of \Eqref{eq:HRK} (up to a constant), and the RK state is the ground state for all values of $t$.

@@

We may understand the emergence of the $Z_2$ topologically ordered phase of IGTs from the QDM and QFPLM from the following picture. The additional vertex configurations that are allowed in the IGTs relative to the models with a dimer number constraint  have U(1) charge $Q_v = \pm 2$ (see \tabref{tab:defects}). It is known that coupling a $U(1)$ gauge field to a dynamic charge $q$ matter field can result in a $Z_q$ gauge theory~\cite{Fradkin1979}; as in the IGT's we have added dynamic charge 2 defects to the underlying $U(1)$ symmetric model, and we may expect a $Z_2$ gauge theory to result. We note that pentamers are charge $\pm3$ objects in the QDM (see \tabref{tab:defects}); thus below we investigate whether introducing pentamers into the QDM leads to an effective $Z_3$ gauge theory (see \appref{sec:Z3GT} for a review of $Z_3$ gauge theory).

@@

We will now present a numerical study of the dimer density correlations in the RK ground state of the QDPM. In \figref{fig:dimer_3way} (a) we plot the bare dimer-dimer density correlation function $\langle n_{\bm{0}} n_{\bm{r}}\rangle$ on a $L=32$ lattice; here the origin $\bm{0}$ is represented by the orange link, and $\bm{r}$ varies over all other links of the lattice. These bare dimer correlations display only short range correlations with no obvious symmetry breaking. In \figref{fig:dimer_3way} (b) we show the equal-time dimer-dimer structure factor, 
\begin{equation}
S_d\left( \bm{k} \right) \equiv \frac{1}{ L^4} \sum_{i,j} e^{-i \bm{k} \cdot \left(\bm{r}_i - \bm{r}_j\right)} \left \langle n_{\bm{r}_i} n_{\bm{r}_j} \right \rangle
\end{equation}
where we restrict the sum to horizontal links, for a $L=64$ lattice. 
\figref{fig:dimer_3way} (c) shows $S_d(\bm{k})$ for a particular path through the Brillouin zone. The dimer structure factor displays broad maxima at $\bm{k} = (0, \pm \pi)$ and $\bm{k} = (\pm \pi, \pm \pi)$. The lack of rotational symmetry of $S_d(\bm{k})$ is due to the fact that we have isolated horizontal dimers; the correlations between vertical dimers display the same behavior, rotated by $\pi/2$.
We have plotted the finite size scaling of the intensity of the broad maximum of $S_d$ in  \figref{fig:bragg_scaling}, which we find to vanish in the thermodynamic limit; these maxima are therefore consistent with finite size effects due to short distance correlations. In \appref{sec:dimer_order_parameters} we additionally have plotted order parameters that specifically test for rotational and translational symmetry breaking; in all cases we find no symmetry breaking in the thermodynamic limit.

@@

Additionally we have studied the nature of the decay of the connected dimer-dimer correlation function,
  \begin{equation}
    C_{\mathrm{d}} \left(r\right) \equiv \langle n_{\bm{0}} n_{\bm{r}} \rangle - \langle n_{\bm{0}} \rangle   \langle n_{\bm{r}} \rangle   
\end{equation}
where $r$ is the distance between the links $\bm{0}$ and $\bm{r}$. In \figref{fig:spatial_dmr_cor}, $C_d$ between parallel links is plotted as a function of separation, for $\bm{r}$ perpendicular to the orientation of the dimers for a $L=64$ lattice. We find that $C_d$ decays exponentially with distance having a correlation length  $\chi\approx 1$. Additionally, $C_d$ displays a sub-lattice oscillation that is consistent with the finite-size maximum in $S_d$ at $(0,\pi)$. Thus we find that the RK point of the QDPM is a symmetric dimer liquid with exponentially decaying dimer correlations, in contrast with the power-law decaying correlations of the QDM.
